---
title: IDW using NOAA data
author: Carles Milà
date: '2020-04-05'
slug: global-and-local-idw
tags:
  - Spatial statistics
  - Geostatistics
  - Nonparametric
  - Environment
  - R
---

```{r knitr_global_options, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center',
                      warning=FALSE, message=FALSE)
set.seed(23)
```

# Basics of IDW 

Inverse Distance Weighted (IDW) is a simple, nonparametic interpolation method consisting on a weighted average of the observations based on the distance (typically on a two-dimensional space for spatial processes). It relies on the first law of geography, saying that all is related, but near things are more related than distant things. Formally, it is defined as (Bivand et al, 2008):

\[ \hat{Z}(s_0) = \frac{\sum^n_{i=1} w(s_i) Z(s_i)}{\sum^n_{i=1}w(s_i)} \]

where $Z$ and $\hat{Z}$ are the observed and estimated values of the continuous spatial process, $s$ is a location typically defined in a two-dimensional space $(x,y)$, and $w$ are the weights defined as: 

\[ w(s_i)=||s_i - s_0||^{-p} \]

that is, the Euclidian distance between two locations at the power of $-p$. For $p = 0$, all observations will have the same weight no matter the distance between the data point $s_i$ and the target point $s_0$, while larger values for $p$ will result in giving larger relative weights to the nearest observations. We can tune the value of $p$ using out-of-sample methods such as cross-validation. We can also limit the amount of neighbours to take into consideration to speed up computations.

IDW is intuitive and quick to implement and it will always yield estimated values ranging between $[min(Z_i), max(Z_i)]$. However, we obtain no estimate of the uncertainty associated with the predictions, it offers a limited flexibility in terms of modelling the correlation between observations based on distance, and ignores the spatial disposition of the observations.

I'll illustrate this method using precipitation data for Germany (in mm) extracted from the NOAA API. This is the list of packages I will use:

```{r}
library("rnoaa") # Download meteo data from NOAA
library("rnaturalearth") # Country boundaries
library("gstat") # IDW function
library("tmap")
library("sf")
library("tidyverse")
```

```{r, echo=FALSE}
api_key <- "KuNaGjLQVWdrcnfceqtQovCAlDJdqDud"
```


# Preparing precipitation data 

Precipitation data (in mm) for Germany, 2018 will be obtained using the beautiful [rnoaa package](https://github.com/ropensci/rnoaa) by [ropensci](https://ropensci.org/). An API token can be freely obtained as indicated in the package documentation.

```{r download_data, eval=FALSE}
# Get location indicators for yearly data (GSOY) at the country level and get Germany
locations <- ncdc_locs(datasetid='GSOY', locationcategoryid = "CNTRY",
                       token = api_key, limit = 1000)
locations <- locations$data %>%
  filter(name %in% c("Germany"))
# Get precipitation data
meteodata <- ncdc(datasetid = "GSOY", 
                  datatypeid = "PRCP",
                  locationid = locations$id,
                  startdate = as.Date("2018-01-01"),
                  enddate = as.Date("2018-12-31"),
                  limit = 1000, 
                  token = api_key) %>%
  .$data
# Get station data
stationdata <- map_df(meteodata$station,
                      ~(ncdc_stations(datasetid='GSOY', stationid = .x, token = api_key)$data))
stationdata$station <- meteodata$station
# We merge the two tables
precipdata <- inner_join(meteodata, stationdata, by = "station")
# Rename precipitation variable
precipdata <- rename(precipdata, precipitation = value)
```

```{r read_data, echo=FALSE}
# write_csv(precipdata, path="data/precip_de_2018.csv")
precipdata <- read_csv("data/precip_de_2018.csv")
precipdata <- rename(precipdata, precipitation = value)
```

We want to create geometries with `sf` to be able perform spatial analysis. We also want to get country boundaries, included in the [rnaturalearth](https://ropensci.org/tutorials/rnaturalearth_tutorial/) package. We'll use ETRS/UTM 31N CRS for Germany. 

```{r convert_sf}
# Create geometries
precipdata <- st_as_sf(precipdata, coords = c("longitude", "latitude"), dim = "XY", crs = 4326) %>%
  st_transform(crs = 25831)
# Germany polygon
germany <- ne_countries(country = "germany", scale = "medium") %>%
  st_as_sf() %>%
  st_transform(crs = st_crs(precipdata))
# Subset stations to Germany
precipdata <- precipdata[st_intersects(precipdata, germany, sparse = FALSE),]
```

Now we are ready to do a cartographic representation of the data using the `tmap` package.

```{r explo_map, fig.width=5, fig.height=4}
tm_shape(germany) +
    tm_borders() +
tm_shape(precipdata) +
    tm_dots(col = "precipitation", style = "jenks", size = 0.25, n = 5, palette = "-cividis") +
tm_layout(legend.outside = TRUE)
```

It appears that the NE of the country was much drier than the South in 2018.

# Tuning p in IDW

As mentioned above, $p$ is the parameter that will determine the weights based on the distance between the target location and the observed data. The larger the $p$, the larger the relative weight of nearby observations (with respect to the rest) will be. To illustrate this, let's compute IDW interpolations for a different range of $p$ values.

```{r idw_different_p, results='hide'}
# We make 10km prediction grid
grid_ger <- st_make_grid(germany, cellsize = 10000, what = "centers")
# Function to perform IDW for a given p
idw_p <- function(p){
  idwgrid <- idw(precipitation ~ 1, precipdata, grid_ger, idp = p) %>%
    rename(precipitation = var1.pred) %>%
    select(precipitation) %>% 
    mutate(p = p)
}
# Run IDW for a series of p, then join results
idw_results <- 0:5 %>%
  map(~ idw_p(p = .x))
idw_results <- do.call(what = sf:::rbind.sf, args = idw_results)
```

```{r map_idw_p_results}
# Make nice labels for p
idw_results$plab <- paste0("p = ", idw_results$p)
# Make map with different p values
tm_shape(germany) +
    tm_borders() +
tm_shape(idw_results) +
    tm_dots(col = "precipitation", size = 0.25, palette = "-cividis", style = "cont") +
tm_facets("plab") +
tm_layout(legend.outside = TRUE)
```

As the resulting maps show, the smaller the $p$ parameter, the smoother the resulting surface is, being the result of $p=0$ a constant surface equal to the precipitation average. A usual default for this parameter is 2, although we can do some parameter search to guide the choice. To do so, we can use the in-built functions in `gstat` and calculate `RMSE` comparing the observed and predicted values.

```{r tuning_p, results='hide'}
# Function to cross-validate IDW
cv_idw <- function(p){
  # Define gstat object
  idw.gstat <- gstat(id = "precipitation", formula = precipitation ~ 1, 
                     data = precipdata, set = list(idp = p))
  # Perform CV
  cv <- gstat.cv(object = idw.gstat, nfold = 10)
  # Calculate RMSE
  res <- tibble(p = p, RMSE = sqrt(mean((cv$precipitation.pred - cv$observed)^2)))
  return(res)

}
# Let's cv for a range of p
cv_res <- seq(1, 6, by = 0.1) %>%
  map_df(~ cv_idw(p = .x))
```

We represent the results graphically to see where the minimum RMSE is.

```{r plot_optimal_p, fig.width=6, fig.height=3}
# And plot results
ggplot(cv_res, aes(x = p, y = RMSE)) +
  geom_point() +
  geom_line() +
  geom_point(data = filter(cv_res, RMSE == min(cv_res$RMSE)), colour = "red", size = 4) +
  theme_bw()
```

For these data, the value of $p$ that minimises the RMSE is `r cv_res$p[cv_res$RMSE==min(cv_res$RMSE)]`.
